---
title: "SiRAcon Presentation Outline"
date: '2022-04-24'
output:
  html_notebook:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Outline of my 2022 SiRAcon 2022 presentation, "Making R work for you (with automation!)".

# Questions/TODO

- [ ] Questions/TODO list here

```{r setup, message = FALSE, warning = FALSE}
library(gert)
library(purrr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(jbplot)
```

# Framework

Use the [DORA Research Program](assets/dora_research_program.pdf) to frame the story of how I
learned R and software engineering by implementing the DORA DevOps technical practices:

- [Version Control](https://cloud.google.com/solutions/devops/devops-tech-version-control)
- [Trunk-based Development](https://cloud.google.com/architecture/devops/devops-tech-trunk-based-development)
- [Shifting Left on Security](https://cloud.google.com/architecture/devops/devops-tech-shifting-left-on-security)
- [Continuous Integration](https://cloud.google.com/architecture/devops/devops-tech-continuous-integration)
- [Continuous Testing](https://cloud.google.com/architecture/devops/devops-tech-test-automation)
- [Deployment Automation](https://cloud.google.com/architecture/devops/devops-tech-deployment-automation)
- [Code Maintainability](https://cloud.google.com/architecture/devops/devops-tech-code-maintainability)
  (also cover "clean" code)

<https://www.devops-research.com/research.html>

# Timeline

Create a timeline of my journey using [vistime](https://shosaco.github.io/vistime/) or
[timevis](https://daattali.com/shiny/timevis-demo/).

# Data

Use GitHub data to show how the DORA metrics changed over time as I developed and used
[rdev](https://jabenninghoff.github.io/rdev/).

## git log

Idea: use `gert::git_log()` tables across all my public and personal (private) R repositories over time to create an
annotated timeline and visualization of my work, and implementation of the DORA technical practices.

Import git logs from my repositories:

```{r git_logs}
my_gitlogs <- map_dfr(
  # nolint start: line_length_linter
  c("jbplot", "rdev", "rstudio-training", "rtraining", "siracon2022", "software-resilience", "workshop7"),
  ~ mutate(git_log(repo = paste0("../../", .x), max = .Machine$integer.max), repo = .x, .before = "commit")
  # nolint end
) |>
  arrange(time)
```

Filter logs by repository, adding cutoff dates when active development ended for timeline visualization.

```{r filtered_gitlogs}
filtered_gitlogs <- my_gitlogs |>
  filter(!(repo == "rstudio-training" & time > ymd_h("2020-12-29 0"))) |>
  filter(!(repo == "software-resilience" & time > ymd_h("2021-02-22 0"))) |>
  filter(!(repo == "rtraining" & time > ymd_h("2021-10-08 0"))) |>
  filter(!(repo == "workshop7" & time > ymd_h("2021-12-07 0"))) |>
  filter(!(repo == "jbplot" & time > ymd_h("2022-02-07 0"))) |>
  mutate(repo = factor(repo, levels = c(
    # nolint start: line_length_linter
    "rstudio-training", "software-resilience", "rtraining", "rdev", "workshop7", "jbplot", "siracon2022"
    # nolint end
  )))
```

Plot monthly commits by repository.

```{r plot_gitlogs}
filtered_gitlogs |>
  mutate(time = floor_date(time, unit = "month")) |>
  group_by(time, repo) |>
  summarize(commits = n(), .groups = "drop") |>
  ggplot(aes(x = time, y = commits, color = repo)) +
  geom_point() +
  geom_line() +
  theme_quo()
```

## Timeline

Create a timeline using groups showing the history of the repositories:

- training: rstudio-training, rtraining, "Working with R"
- development: rdev, jbplot
- notebooks: software-resilience, workshop7, siracon2022

Key dates:

2020-09-08: Starting out, rstudio-training, renv

- Version Control
  - Stored project files, notebook (Rmd and html) in private git repository
  - Use renv to store package dependencies in source control
- Trunk-based Development
  - Direct commits to master (not recommended)

2020-09-30: (Aside) First bug discovered, https://github.com/rstudio/renv/issues/547 !

2020-10-06: setup-r script

- Version Control
  - Automate setup of local R development environment

2020-12-02: Adoption of styler and lintr

- Code Maintainability
  - Consistent formatting (styler)
  - Consistent code (lintr)
- Continuous Testing
  - Static code analysis (lintr)

2020-12-27: Migration to rtraining package

- Continuous Integration
  - Build Package
- Continuous Testing
  - R CMD Check
- Version Control
  - .Rprofile

2020-12-29: build-site script

- Deployment Automation
  - build-site: shell script to publish notebooks using `rmarkdown::render_site()`
  - MVP for publishing notebooks using GitHub Pages

## Releases

Plot releases over time: total GitHub releases per period (for all repositories) to show changes in release frequency.
