---
title: "SiRAcon Presentation Outline"
date: '2022-04-24'
output:
  html_notebook:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Outline of my 2022 SiRAcon 2022 presentation, "Making R work for you (with automation!)".

# Questions/TODO

- [ ] Questions/TODO list here

```{r setup, message = FALSE, warning = FALSE}
library(siracon2022)
library(readr)
library(gert)
library(gh)
library(purrr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(vistime)
library(timevis)
library(jbplot)
```

# Framework

Use the [DORA Research Program](assets/dora_research_program.pdf) to frame the story of how I
learned R and software engineering by implementing the DORA DevOps technical practices:

- [Version Control](https://cloud.google.com/solutions/devops/devops-tech-version-control)
- [Trunk-based Development](https://cloud.google.com/architecture/devops/devops-tech-trunk-based-development)
- [Shifting Left on Security](https://cloud.google.com/architecture/devops/devops-tech-shifting-left-on-security)
- [Continuous Integration](https://cloud.google.com/architecture/devops/devops-tech-continuous-integration)
- [Continuous Testing](https://cloud.google.com/architecture/devops/devops-tech-test-automation)
- [Deployment Automation](https://cloud.google.com/architecture/devops/devops-tech-deployment-automation)
- [Code Maintainability](https://cloud.google.com/architecture/devops/devops-tech-code-maintainability)
  (also cover "clean" code)

<https://www.devops-research.com/research.html>

Create a timeline of my journey using [vistime](https://shosaco.github.io/vistime/) or
[timevis](https://daattali.com/shiny/timevis-demo/).

# Outline

Use GitHub data to show how the DORA metrics changed over time as I developed and used
[rdev](https://jabenninghoff.github.io/rdev/).

## git log

Idea: use `gert::git_log()` tables across all my public and personal (private) R repositories over
time to create an annotated timeline and visualization of my work, and implementation of the DORA
technical practices.

Import git logs from my repositories:

```{r gitlogs}
# TODO: add data to package using methods from https://r-pkgs.org/data.html
# TODO: open issue on gert: git_log shows rebase & merge time, while "git log" shows original time
gitlogs <- map_dfr(
  c(
    "jbplot", "rdev", "rstudio-training", "rtraining", "siracon2022", "software-resilience",
    "workshop7"
  ),
  ~ mutate(git_log(repo = paste0("~/GitHub/", .x), max = .Machine$integer.max),
    repo = .x, .before = "commit"
  )
) |>
  arrange(time)

gitlogs_tz <- tz(gitlogs$time)
```

Filter logs by repository, adding cutoff dates when active development ended for timeline
visualization. Drop commits past April 30 to remove partial months.

```{r filtered_gitlogs}
filtered_gitlogs <- gitlogs |>
  # set filter to midnight after last relevant commit, use same timezone as gitlogs
  filter(!(repo == "rstudio-training" & time > ymd_h("2020-12-28 0", tz = gitlogs_tz))) |>
  filter(!(repo == "software-resilience" & time > ymd_h("2021-02-22 0", tz = gitlogs_tz))) |>
  filter(!(repo == "rtraining" & time > ymd_h("2021-10-08 0", tz = gitlogs_tz))) |>
  filter(!(repo == "workshop7" & time > ymd_h("2021-12-07 0", tz = gitlogs_tz))) |>
  filter(!(repo == "jbplot" & time > ymd_h("2022-02-07 0", tz = gitlogs_tz))) |>
  filter(time < ymd_h("2022-05-01 0", tz = gitlogs_tz)) |>
  # oldest first
  mutate(repo = factor(repo, levels = c(
    "rstudio-training", "software-resilience", "rtraining", "rdev", "workshop7", "jbplot",
    "siracon2022"
  )))
```

Plot monthly commits by repository.

```{r monthly_commits_repo}
filtered_gitlogs |>
  mutate(time = floor_date(time, unit = "month")) |>
  group_by(time, repo) |>
  summarize(commits = n(), .groups = "drop") |>
  ggplot(aes(x = time, y = commits, color = repo)) +
  geom_point() +
  geom_line() +
  labs(title = "Monthly commits by repository") +
  labs(x = "", y = "", color = "repository") +
  theme_quo()

ggsave("rendered/monthly-commits-repo.png", width = 16 * 0.6, height = 9 * 0.6, bg = "white")
```

[High resolution plot](rendered/monthly-commits-repo.png)

## Timeline

Create a timeline using groups showing the history of the repositories:

- training: rstudio-training, rtraining
- development: rdev, jbplot
- notebooks: software-resilience, workshop7, siracon2022

```{r repo_timeline}
repo_timeline <- filtered_gitlogs |>
  select(repo, time) |>
  mutate(time = floor_date(time, unit = "day")) |>
  group_by(repo) |>
  summarize(start = min(time), end = max(time)) |>
  arrange(start) |>
  mutate(group = case_when(
    grepl("training", repo) ~ "training",
    repo %in% c("rdev", "jbplot") ~ "development",
    TRUE ~ "notebooks"
  )) |>
  mutate(color = hue_pal()(7)[row_number()])

repo_timeline
# TODO: issue: gg_vistime doesn't render well when using scale_color_viridis_d()
gg_vistime(repo_timeline, col.event = "repo", title = "R Development Timeline") +
  theme_quo()

ggsave("rendered/repo-timeline.png", width = 16 * 0.6, height = 9 * 0.6, bg = "white")
```

[High resolution plot](rendered/repo-timeline.png)

## Key events

Plot key events on a `timevis()` timeline. [Full page version](rendered/key-events.html).

```{r key_events, fig.width = 10, fig.height = 15}
key_events <- read_csv("data/key-events.csv", col_types = cols(
  id = col_integer(),
  start = col_date(format = ""),
  end = col_date(format = ""),
  content = col_character(),
  group = col_integer(),
  group_content = col_character(),
  intro = col_logical(),
  milestone = col_logical()
))

dora_groups <- key_events |>
  select(id = group, content = group_content) |>
  unique() |>
  arrange(id)

key_events |>
  render_timevis(groups = dora_groups, file = "rendered/key-events.html", showZoom = TRUE)
```

2020-09-08: Starting out, rstudio-training, renv

- Version Control
  - Stored project files, notebook (Rmd and html) in private git repository
  - Use renv to store package dependencies in source control
- Trunk-based Development
  - Direct commits to master (not recommended)
- Shifting Left on Security
  - Start development with `renv::update()`

2020-09-11: Published ["Working with R"](https://www.information-safety.org/2020/09/11/working-with-r/)

2020-09-30: (Aside) First bug discovered, https://github.com/rstudio/renv/issues/547 !

2020-10-06: `setup-r` script

- Version Control
  - Automate setup of local R development environment

2020-12-02: Adoption of styler and lintr

- Code Maintainability
  - Consistent formatting (styler)
  - Consistent code (lintr)
- Continuous Testing
  - Static code analysis (lintr)

2020-12-27: Migration to rtraining package

- Continuous Integration
  - Build Package
- Continuous Testing
  - R CMD check
- Version Control
  - .Rprofile

2020-12-29: build-site script

- Deployment Automation
  - build-site: shell script to publish notebooks using `rmarkdown::render_site()`
  - MVP for publishing notebooks using GitHub Pages

2020-12-30: First release: rtraining 0.0.1

2020-12-30: GitHub Actions

- Continuous Integration, Continuous Testing
  - R-CMD-check
  - lintr

2020-12-30: `lint_all()`

- Continuous Integration, Continuous Testing
  - lint all files locally
  - first testthat tests
  - roxygen2 documentation

2020-12-30: `style_all()`

- Continuous Integration, Code Maintainability
  - run styler on all files locally

2020-12-31: Switch GitHub Actions to `lint_all()`

- Continuous Integration, Continuous Testing
  - match GitHub and local CI checks

2021-01-01: `ci()`, `check_renv()`

- Continuous Integration, Continuous Testing
  - run all CI checks locally
  - eliminate toil
  - match GitHub and local CI checks

2021-01-01: Migration to rdev package

- Code Maintainability
  - Moved most functions to new rdev package
  - Consistent tools across projects

2021-01-02: Multi-platform R CMD check

- Continuous Integration, Continuous Testing
  - ensure package works on Windows and macOS

2021-01-03: First version of `build_analysis_site()`

- Deployment Automation
  - Automatically build GitHub Pages site with functions, notebooks
  - Still a shell script
  - Beginning of standard deployment and release pattern:
    - bump version
    - write code
    - update NEWS.md
    - "GitHub Release"
    - build_site

2021-01-09: Analysis Package Layout

- Code Maintainability
  - Consistent package layout across projects
  - Supported future automation for creating packages

2021-01-12: Native R version of `build_analysis_site()`

2021-01-16: Migrated `build_analysis_site()` from rtraining to rdev

- Code Maintainability
  - Cross-platform support
  - Moves all automation to R Console
- Deployment Automation
  - Automated builds across all projects

2021-09-29: Formal R Analysis Package Layout, Documented release process

- Code Maintainability
  - Consistent package layout across projects
  - Supported future automation for creating packages
- Deployment Automation
  - Supported future automation for creating releases

2021-12-04: Documented package creation process

- Code Maintainability
  - Consistent package layout across projects
  - Supported future automation for creating packages

2021-12-23: `theme_quo()`: a personalized theme to visually identify my ggplots.

2022-01-01: Automate package configuration with `use_analysis_package()`

- Code Maintainability
  - Consistent package layout across projects

2022-01-10: Create package automation (rdev 0.7.0)

- `create_github_repo()`: Create new GitHub repository following rdev conventions in the active
  user's account and create a basic package
- `use_rdev_package()`: Add rdev templates and settings within the active package. Normally invoked
  when first setting up a package.
- Added `build_rdev_site()`, a wrapper for `pkgdown::build_site()` optimized for rdev workflow that
  updates `README.md` and performs a clean build using pkgdown
- Added 'Analysis Notebook' R markdown template for RStudio (File > New File > Rmarkdown > From
  Template)
- Migrated ggplot2 themes/styles (`theme_quo()`, `viridis_quo()`) to new package,
  `jabenninghoff/jbplot`

- Code Maintainability
  - Cross-platform support
  - Moves all automation to R Console
- Deployment Automation
  - Automated builds across all projects

2022-01-10: Automate notebook listings in README

- Deployment Automation

2022-01-17: Release automation (rdev 0.8.0)

- `stage_release()`: Open a GitHub pull request for a new release from NEWS.md
- `merge_release()`: Merge a staged pull request and create a new GitHub release

- Deployment Automation

2022-01-19: More workflow automation

- Added `new_branch()`: Create a new feature branch, and (optionally) bump the version in
  DESCRIPTION

2022-01-21 - 2022-02-06: adding test coverage

- Continuous Testing
  - Biggest challenge yet
  - Significantly improved code quality
  - "Unit" testing
  - Just test
  - Test program flow
  - Don't test other people's code
  - Mock external functions
  - Fix bugs by writing a test
  - Code coverage, and code coverage metrics
  - Test Driven Development
  - Tests Give You Confidence (to Refactor)

(Show plot of increasing code coverage from codecov.io)

2022-01-24: write_eval() is a really bad idea:

```r
#' Write and evaluate an expression
#'
#' `write_eval(string)` is a simple wrapper that prints `string` to the console using
#'   [`writeLines()`][base::writeLines], then executes the expression using [`parse()`][base::parse]
#'   and [`eval()`][base::eval].
#'
#' @param string An expression to be printed to the console and evaluated
#'
#' @return The return value from the evaluated expression
#'
#' @examples
#' write_eval("pi")
#'
#' write_eval("exp(1)")
#' @export
write_eval <- function(string) {
  if (!is.character(string)) stop("not a character vector")
  if (string == "") stop("nothing to evaluate")
  writeLines(string)
  eval(parse(text = string))
}
```

2022-01-30: Manual test script for new package setup

- Continuous Testing
  - Manual tests evolve into partially or fully automated tests

2022-02-02: Added `local_temppkg()` test helper function

- Continuous Testing
  - Test helpers - testing test helpers helps!

2022-02-06: rdev 1.0.0 !

- Release automation: Stage and create GitHub releases, including GitHub pages
- Continuous Integration: Local continuous integration checks and dependency management
- Package Setup: Package setup tasks, typically performed once

2022-02-06 - Today: Continuous Improvement

- Improve CI workflow to catch mistakes
- Spell checks
- Branch protection automation
- Options
- Dependency management
- Product health

## Releases

Get releases from GitHub using `siracon2022::gh_releases()`:

```{r releases}
# cache results
if (!exists("releases")) {
  releases <- map_dfr(
    c("rtraining", "rdev", "workshop7", "jbplot", "siracon2022") %>% setNames(., .),
    gh_releases, "jabenninghoff",
    .id = "repo"
  ) |>
    arrange(time)
}
```

Filter releases past April 30 to remove partial months.

```{r filtered_releases}
filtered_releases <- releases |>
  mutate(time = with_tz(time, tzone = gitlogs_tz)) |>
  filter(time < ymd_h("2022-05-01 0", tz = gitlogs_tz))
```

Plot releases over time: total GitHub releases per period (for all repositories) to show changes in
release frequency. The dotted line marks the implementation of release automation.

```{r monthly_releases}
monthly_releases <- filtered_releases |>
  mutate(time = floor_date(time, unit = "month")) |>
  group_by(time) |>
  summarize(releases = n(), .groups = "drop") |>
  add_row(time = ymd("2020-11-01"), releases = 0) |>
  add_row(time = ymd("2020-10-01"), releases = 0) |>
  add_row(time = ymd("2020-09-01"), releases = 0) |>
  arrange(time)

monthly_releases |>
  ggplot(aes(x = time, y = releases)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = ymd_h("2020-12-01 0", tz = gitlogs_tz), linetype = "dotted") +
  geom_vline(xintercept = ymd_h("2022-01-01 0", tz = gitlogs_tz), linetype = "dotted") +
  coord_cartesian(ylim = c(0, NA)) +
  labs(title = "Monthly GitHub releases") +
  labs(x = "", y = "") +
  theme_quo()

ggsave("rendered/monthly-releases.png", width = 16 * 0.6, height = 9 * 0.6, bg = "white")
```

[High resolution plot](rendered/monthly-releases.png)

However, the number of releases per month might just represent how much work is being done, and
looks similar to the plot of all commits by month:

```{r monthly_commits}
gitlogs |>
  filter(time < ymd_h("2022-05-01 0", tz = gitlogs_tz)) |>
  mutate(time = floor_date(time, unit = "month")) |>
  group_by(time) |>
  summarize(commits = n(), .groups = "drop") |>
  arrange(time) |>
  ggplot(aes(x = time, y = commits)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = ymd_h("2020-12-01 0", tz = gitlogs_tz), linetype = "dotted") +
  geom_vline(xintercept = ymd_h("2022-01-01 0", tz = gitlogs_tz), linetype = "dotted") +
  coord_cartesian(ylim = c(0, NA)) +
  labs(title = "Monthly git commits") +
  labs(x = "", y = "") +
  theme_quo()

ggsave("rendered/monthly-commits.png", width = 16 * 0.6, height = 9 * 0.6, bg = "white")
```

[High resolution plot](rendered/monthly-commits.png)

Also plot releases per commit, which will fall between 0 and 1. The dotted lines mark adoption of
GitHub and implementation of release automation.

```{r releases_per_commit}
gitlogs |>
  filter(time < ymd_h("2022-05-01 0", tz = gitlogs_tz)) |>
  mutate(time = floor_date(time, unit = "month")) |>
  group_by(time) |>
  summarize(commits = n()) |>
  full_join(monthly_releases, by = "time") |>
  replace_na(list(commits = 0, releases = 0)) |>
  mutate(rpc = releases / commits) |>
  ggplot(aes(x = time, y = rpc)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = ymd_h("2020-12-01 0", tz = gitlogs_tz), linetype = "dotted") +
  geom_vline(xintercept = ymd_h("2022-01-01 0", tz = gitlogs_tz), linetype = "dotted") +
  labs(title = "Monthly GitHub releases per commit") +
  labs(x = "", y = "") +
  theme_quo()

ggsave("rendered/releases-per-commit.png", width = 16 * 0.6, height = 9 * 0.6, bg = "white")
```

[High resolution plot](rendered/releases-per-commit.png)

# Story

Use the timeline and plots to tell the story of continuous improvement. Each section filters on
group 1 and the other focus area. Integrate themes into story.

1. Introduction: background and motivation, use Event group as the talk overview. Exclude SiRAcon
   2020 from future timelines. "R Development Timeline".
1. Version Control: put everything (except artifacts) into version control for reproducibility and
   history.
1. Trunk-based Development: linear development avoids code conflicts.
1. Shift Left on Security: maintenance first ensures you get it done.
1. Continuous Integration: build and test on each commit to catch mistakes early.
1. Deployment Automation: automate your development workflow to spend more time writing.
1. Code Maintainability: consistent and clean code is easier to understand.
1. Continuous Testing: (the biggest challenge) formally specifying what you are building and how it
   is *supposed* to work defends against the dangers of hidden assumptions.
1. Results: "Monthly commits by repository", "Monthly GitHub releases", "GitHub releases per
   commit". Improvement on technical practices also means less rework, less deployment pain, less
   burnout, and greater job satisfaction.
1. Closing: complete key events timeline.

Full rdev package list:

- `desc`
- `devtools`
- `fs`
- `gert`
- `gh`
- `lintr`
- `markdown`
- `miniUI`
- `pkgdown`
- `purrr`
- `rcmdcheck`
- `remotes`
- `renv`
- `rlang`
- `rmarkdown`
- `styler`
- `tibble`
- `usethis`
- `withr`
- `xml2`
- `yaml`
- `covr`
- `DT`
- `knitr`
- `mockery`
- `spelling`
- `stringi`
- `testthat`

## Introduction

Background and motivation. [Full page version](rendered/intro.html).

```{r intro, fig.width = 8.6, fig.height = 3.5}
key_events |>
  filter(group == 1) |>
  render_timevis(groups = filter(dora_groups, id == 1), file = "rendered/intro.html")
```

## Version Control

Put everything (except artifacts) into version control for reproducibility and history.
[Full page version](rendered/version-control.html).

*Use of [Homebrew](https://brew.sh), and `brew bundle`.*

Packages:

- `renv`: dependency management

```{r version_control, fig.width = 8.6, fig.height = 2.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 2) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 2)), "rendered/version-control.html")
```

## Trunk-based Development

Linear development avoids code conflicts. [Full page version](rendered/trunk-based.html).

```{r trunk_based, fig.width = 8.6, fig.height = 2.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 3 | id == 44) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 3)), file = "rendered/trunk-based.html")
```

## Shift Left on Security

Maintenance first ensures you get it done. [Full page version](rendered/shift-left.html).

*Reference last year's talk, recording available in member's section.*

Packages:

- `renv`

```{r shift_left, fig.width = 8.6, fig.height = 2.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 4) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 4)), file = "rendered/shift-left.html")
```

## Continuous Integration

Build and test on each commit to catch mistakes early. [Full page version](rendered/ci.html).

Packages:

- `devtools`
- `usethis`
- [r-lib/actions](https://github.com/r-lib/actions)

```{r ci, fig.width = 8.6, fig.height = 3.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 5) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 5)), file = "rendered/ci.html")
```

## Deployment Automation

Automate your development workflow to spend more time writing.
[Full page version](rendered/deployments.html).

*Reducing toil. Forming habits, which become repeated tasks, which become automation. If it's
automated, it gets done.*

Packages:

- `pkgdown`, `rmarkdown`: `build_analysis_site()`
- `gert`, `gh`: git, GitHub automation
- `devtools`

```{r deployments, fig.width = 8.6, fig.height = 4.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 6 | id == 32) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 6)), file = "rendered/deployments.html")
```

## Code Maintainability

Consistent and clean code is easier to understand. [Full page version](rendered/code-maint.html).

*Functional programming (purrr) vs procedural programming. Functional programming is harder to
learn, but safer.*

*R dialects: base R is for functions, tidyverse R is for notebooks.*

*"Clean" code: code should be written for future humans, including you!*

Packages:

- `styler`
- `usethis`
- `gh`
- `desc`
- `devtools`
- `purrr`

```{r code_maint, fig.width = 8.6, fig.height = 4}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 7 | id == 32) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 7)), file = "rendered/code-maint.html")
```

## Continuous Testing

The biggest challenge: formally specifying what you are building and how it is *supposed* to work
defends against the dangers of hidden assumptions. [Full page version](rendered/testing.html).

Packages:

- `lintr`: static code analysis
- `rcmdcheck`
- `testthat`
- `devtools`
- `covr`
- `mockery`
- `withr`
- `rlang`
- `spelling`

### Future Testing

Mutation Testing: [Wikipedia](https://en.wikipedia.org/wiki/Mutation_testing)

R packages:

- [mutant](https://github.com/sckott/mutant)
- [autotest](https://github.com/ropensci-review-tools/autotest/)

Papers:

- [Does mutation testing improve testing practices?](https://homes.cs.washington.edu/~rjust/publ/mutation_testing_practices_icse_2021.pdf)
- [Practical Mutation Testing at Scale](https://homes.cs.washington.edu/~rjust/publ/practical_mutation_testing_tr_2021.pdf)

Formal Methods:

- [Planning with flare](https://increment.com/planning/formal-specifications-and-planning/)
- [Hillel Wayne](https://hillelwayne.com)
- [Learn TLA+](https://learntla.com/introduction/)
- [Alloy Documentation](https://alloy.readthedocs.io/en/latest/)

```{r testing, fig.width = 8.6, fig.height = 3.5}
key_events |>
  filter(!intro) |>
  filter(milestone | group == 8 | id == 32) |>
  render_timevis(groups = filter(dora_groups, id %in% c(1, 8)), file = "rendered/testing.html")
```

End of (out)line.
